\documentclass[8pt]{beamer}

\usepackage{cmap}
\usepackage[T2A]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{color}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{pdfpages}
\usepackage{forest}
\usepackage{adjustbox}
\usepackage{times}
\usepackage{tikz}

\mode<presentation>{
    \usetheme{Marburg}
    \usecolortheme{sidebartab}
}

% \newtheorem{theorem}{Theorem}[section]
% \newtheorem{lemma}{Lemma}[section]
% \newtheorem{proposition}{Proposition}[section]
% \newtheorem{corollary}{Corollary}[section]
% \newtheorem{definition}{Definition}[section]
% \newtheorem{remark}{Remark}[section]
% \newtheorem{example}{Example}[section]

\newcommand{\E}{\ensuremath{\mathbb{E}}}
\newcommand{\D}{\ensuremath{\mathbb{D}}}
\renewcommand{\C}{\ensuremath{\mathbb{C}}}
\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\Q}{\ensuremath{\mathbb{Q}}}
\newcommand{\Z}{\ensuremath{\mathbb{Z}}}
\renewcommand{\|}{\ensuremath{\hspace{0.1cm} | \hspace{0.1cm}}}

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\green}[1]{\textcolor{green}{#1}}
\newcommand{\orange}[1]{\textcolor{orange}{#1}}
\newcommand{\teal}[1]{\textcolor{teal}{#1}}
\newcommand{\purple}[1]{\textcolor{purple}{#1}}

\renewcommand{\phi}{\varphi}
\renewcommand{\epsilon}{\varepsilon}
\renewcommand{\le}{\leqslant}
\renewcommand{\ge}{\geqslant}


\begin{document}
    \title[Liouville's Theorem]{Liouville's Theorem \\ on integrability via elementary functions}
    \author{Vanya Vorobiov}
    \date{\today}
    \institute{Sber}

    \begin{frame}
        \titlepage
    \end{frame}

    \section{Introduction}
    \begin{frame}{Introduction}
        \uncover<1->{
            From high school, we are familiar with the idea that some integrals cannot be expressed in terms of elementary functions. For instance:
            \[
                \int e^{\pm x^2} \, dx, \quad \int \frac{dx}{\ln x}, \quad \int \frac{e^x}{x} \, dx, \quad \int \frac{\sin x}{x} \, dx, \quad \int \frac{\sinh x}{x} \, dx, \quad \int \log \log x \, dx
            \]

            These examples highlight the limitations of elementary functions in representing certain integrals.
        }
        
        \uncover<2->{
            In this presentation, we will:
            \begin{itemize}
                \item Introduce and prove a powerful tool: \textbf{Liouville's theorem}.
                \item Derive these integrals as a consequence.
                \item If time permits, discuss a special \textbf{class} of \textbf{elliptic integrals}, which also cannot be expressed in elementary terms:
                \[
                \int \frac{dx}{\sqrt{P(x)}}
                \]
                where $\deg P = 2, 3$ and $P$ hasn't multiple roots.
            \end{itemize}
        }
    \end{frame}

    \section{Basic definitions}
    \begin{frame}{Basic definitions}
        \begin{block}{Remark}<1->
            Throughout the presentation, we will assume that all fields have characteristic 0.
        \end{block}
        \begin{definition}<2->
            Field $F$ is differential if it's equipped with the unary function $'$ such that:
            \begin{itemize}
                \item $(a+b)' = a' + b'$
                \item $(ab)' = a'b + ab'$
            \end{itemize}
        \end{definition}
        \begin{definition}<3->
            The subfield $K \subseteq F$, defined as $K = { a \in F \mid a’ = 0 }$, is called the subfield of constants.
        \end{definition}
        \begin{definition}<4->
            A differential extension of a differential field $F$ is a field $E$ such that $E \supseteq F$ and the differentiation $’$ extends to $E$ in the same way”.
        \end{definition}
        \begin{definition}<5->
            Let $F$ be a differential field. Then
            \begin{itemize}
                \item $b$ is called the logarithm of $a$ if $b' = \frac{a'}{a}$
                \item $b$ is called the exponent of $a$ if $a' = \frac{b'}{b}$
            \end{itemize}
        \end{definition}
    \end{frame}
    
    

    \section{What is an expression in elementary terms?}
    \begin{frame}{What is an expression in elementary terms?}
        \begin{definition}<1->
            An extension $E$ of $F$ is called \textbf{elementary} if it can be written as $E = F(t_1, \ldots, t_n)$, where for each $i$, the element $t_i$ is either a logarithm, an exponent, or algebraic over $F(t_1, \ldots, t_{i-1})$.
        \end{definition}
        \begin{block}{Remark}<2->
            Common sense tells us that a function $f: \C \to \C$ is elementary if and only if it can be constructed using a finite number of radicals, sines, cosines, exponents, logarithms, and hyperbolic functions.
            One can see that this is consistent with our approach. Furthermore, our definition of elementarity is more general.
        \end{block}
    \end{frame}

    \begin{section}{Liouville's Theorem (statement)}
    \begin{frame}{Liouville's Theorem (statement)}
        \begin{theorem}[Liouville, 1833–1841]
            Let $F$ be a differential field, and let $K$ be its subfield of constants.  
            If for some $\alpha \in F$, the equation $x' = \alpha$ has a solution in some elementary extension of $F$ such that its subfield of constants remains $K$, then  
            \[
                \alpha = \sum\limits_{i=1}^m c_i \frac{u_i'}{u_i} + v'
            \]
            for some $c_1, \ldots, c_m \in K$ and $u_1, \ldots, u_m, v \in F$.
        \end{theorem}
    \end{frame}

    \section{The Main Lemma}
    \begin{frame}{The Main Lemma}
        \begin{lemma}<1->
            Let $F$ be a differential field, and let $t$ be transcendental over $F$. Suppose that $t$ is either a logarithm or an exponent of some element from $F$.  
            Let $f \in F[x]$ be a polynomial with $\deg f = k \geq 1$. Then:
            \begin{itemize}
                \item If $t$ is a logarithm, then the degree of $(f(t))'$ is $k$ if the leading coefficient of $f$ is not a constant, and $k - 1$ if the leading coefficient is a constant.
                \item If $t$ is an exponent, then the degree of $(f(t))'$ is $k$, and it is a multiple of $f$ if and only if $f$ is a monomial.
            \end{itemize}
        \end{lemma}

        \begin{proof}<2->
            This is a straightforward technical exercise.
        \end{proof}
    \end{frame}

    \section{Liouville's Theorem (proof)}
    \begin{frame}{Liouville's Theorem (proof)}
        \uncover<1->{
            Let $x$ be a solution of the differential equation mentioned above, and suppose that $x \in F(t_1, \ldots, t_n)$.

            We will use induction on $n$ (without fixing the field $F$).

            For brevity, we denote $t = t_1$.
        }

        \uncover<2->{
            Using the inductive assumption, we obtain
            \[
                \alpha = \sum\limits_{i=1}^m c_i \frac{u_i'}{u_i} + v'
            \]
            for some $c_1, \ldots, c_m \in K$ and $u_1, \ldots, u_m, v \in F(t)$.
            
            Here we use the fact that the subfield of constants of $F(t)$ is $K$.
        }

        \uncover<3->{
            Now we consider three cases:
            \begin{itemize}
                \item $t$ is transcendental over $F$ and is a logarithm;
                \item $t$ is transcendental over $F$ and is an exponent;
                \item $t$ is algebraic over $F$.
            \end{itemize}
        }
    \end{frame}

    \subsection{$t$ is a trancendental logarithm}
    \begin{frame}{$t$ is a transcendental logarithm}
        \uncover<1->{
            First, let us consider the basic properties of logarithmic derivatives:
            \[
            \frac{(ab)'}{ab} = \frac{a'}{a} + \frac{b'}{b}, \quad \frac{(1/a)'}{1/a} = - \frac{a'}{a}
            \]
        }
        
        \uncover<2->{
            Then we can assume that all $u_i$ are distinct monic irreducible polynomials.
        }

        \uncover<3->{
            Now suppose that some $u_i \notin F$. It is clear that $\frac{u_i'}{u_i}$ is already in lowest terms (since $\deg u_i > \deg u_i'$).
        }

        
        \begin{itemize}
            \uncover<4->{
                \item If there is no $u_i$ in the denominator of $v$, then there is no $u_i$ in the denominator of $v'$. But then $\alpha \notin F$.
            }
            \uncover<5->{
                \item If there is $u_i$ in the denominator of $v$, then the denominator of $v'$ is divisible by $u_i^2$ and still cannot be reduced in the general sum, implying that $\alpha \notin F$.
            }
        \end{itemize}

        \uncover<6->{
            Therefore, $u_1, \ldots, u_m \in F$ and $v' \in F$.
        }

        \uncover<7->{
            Then 
            \[
                v' = (ct + s)' = ct' + s' = c \frac{z'}{z} + s'
            \]
            \[
                \alpha = \sum c_i \frac{u_i'}{u_i} + c \frac{z'}{z} + s'
            \]
        QED.
        }
    \end{frame}

    \subsection{$t$ is a trancendental exponent}
    \begin{frame}{$t$ is a transcendental exponent}
        \uncover<1->{
            Let $u_1, \ldots, u_m$ be distinct monic irreducible polynomials again.
        }

        \uncover<2->{
            It is clear that $u_i = t$ is the only $u$ that may not be in $F$, because $t$ is the only irreducible monomial, and we would get the same contradiction otherwise.
        }

        \uncover<3->{
            By the main lemma, $v'\in F$ if and only if $v\in F$.
        }

        \uncover<4->{
            Then 
            \[
                \alpha = c_1 \frac{t'}{t} + \sum\limits_{i=2}^m \frac{u_i'}{u_i} + v' = \sum\limits_{i=2}^m \frac{u_i'}{u_i} + (v + c_1 z)'
            \]
        }
    \end{frame}

    \subsection{$t$ is algebraic}
    \begin{frame}{$t$ is algebraic}
        \uncover<1->{
            Consider all conjugate elements of $t$:
            \[
                \tau_1=t, \tau_2, \ldots, \tau_k
            \]
        }

        \uncover<2->{
            Then it is clear that for all $j$:
            \[
                \alpha = \sum\limits_i \frac{u_i'(\tau_j)}{u_i(\tau_j)} + v'(\tau_j)
            \]
        }

        \uncover<3->{
            \[
                \alpha = \frac{1}{k} \left( \sum\limits_i \sum\limits_j \frac{u_i'(\tau_j)}{u_i(\tau_j)} + \sum\limits_j v'(\tau_j) \right)
            \]
        }

        \uncover<4->{
            \[
                \alpha = \frac{1}{k} \left( \sum\limits_i \frac{\left(\prod\limits_j u_i(\tau_j)\right)'}{\prod\limits_j u_i(\tau_j)} + \left(\sum\limits_j v(\tau_j)\right)' \right)
            \]
            Since all rational symmetric functions of a polynomial's roots belong to the base field, the proof is complete.
        }

        \uncover<5->{
            \begin{block}{Remark}
                Here we used the fact that an algebraic extension can be equipped with a unique differentiation, but this is trivial in the case of meromorphic functions.
            \end{block}
        }
    \end{frame}

    \section{Corollaries}
    \subsection{The main corollary}
    \begin{frame}{The main corollary}
        \begin{block}{Corollary}<1->
            Let $f, g \in \C(x), f \neq 0$ and $g$ is not constant. 
            If the integral of $f(x) \exp(g(x))$ can be expressed in elementary terms, then there exists $r \in \C(x)$ such that $f = r' + rg'$.
        \end{block}
        \begin{proof}<2->
            \uncover<2->{
                Denote $F = \C(x)$ and $t = \exp(g)$.
            }
            
            \uncover<3->{
                Using Liouville's theorem, we get
                \[
                    ft = \sum c_i \frac{u_i'}{u_i} + v'
                \]
            }
            
            \uncover<4->{
                One can use the main lemma and obtain:
                \[
                    c_i \frac{u_i'}{u_i} \in F, \quad v = \sum\limits_{i \in I \subseteq \Z} b_i t^i
                \]
                
                The rest of the proof is trivial.
            }
        \end{proof}
    \end{frame}
    
    \subsection{Some special cases}
    \begin{frame}{Some special cases}
        \begin{block}{Corollary}<1->
            The following integrals are not elementary:
            \[
                \int e^{\pm x^2} \, dx, \quad \int \frac{dx}{\ln x}, \quad \int \frac{e^x}{x} \, dx, \quad \int \log \log x \, dx
            \]
        \end{block}
        
        \begin{proof}<2->
            All of these cases can be simply proven using the main corollary and some basic asymptotic analysis.
        \end{proof}

        \begin{block}{Corollary}<3->
            The following integrals are not elementary:
            \[
                \int \frac{\sin x}{x} \, dx, \quad \int \frac{\sinh x}{x} \, dx
            \]
        \end{block}

        \begin{proof}<4->

            \uncover<4->{
                % This case is slightly more complicated.
                
                Denote $F = \C(x)$, $t = \exp(x)$, and using Liouville's theorem, we get
                \[
                    \frac{t^2 - 1}{tz} = \sum c_i \frac{u_i'}{u_i} + v'
                \]
            }

            \uncover<5->{
                After some considerations, using the main lemma, we deduce an equation:
                \[
                    \frac{1}{x} = a' + a
                \]
            }
        \end{proof}
    \end{frame}

    \subsection{On elliptic integrals}
    \begin{frame}{On elliptic integrals}
        \begin{block}{Corollary}<1->
            Let $P$ be a monic polynomial of degree $\ge 3$ with no repeated roots. Then
            \[
                \int \frac{dx}{\sqrt{P(x)}} 
            \]
            is not elementary.
        \end{block}

        \begin{proof}<2->
            Here, we will use some more advanced concepts from the theory of Riemann surfaces.

            Liouville's theorem allows us to write
            \[
                \frac{1}{\sqrt{P(x)}} = \sum c_i \frac{u_i'}{u_i} + v'
            \]
            with $c_1, \ldots, c_m \in \C$ and $u_1, \ldots, u_m, v \in \C(x, \sqrt{P(x)})$.
            
            \uncover<3->{
                Now, we deduce an equation on the compact Riemann surface $C$ associated with the equation $y^2 = P(x)$:
                \[
                    \frac{dx}{y} = \sum c_i \frac{du_i}{u_i} + dv
                \]
            }

            \uncover<4->{
                The left-hand side is a nonzero holomorphic $1$-form on $C$.

                However, such forms cannot be expressed as a linear combination of logarithmic meromorphic differentials $\frac{du}{u}$ and exact meromorphic differentials $dv$.
            }
        \end{proof}
    \end{frame}

    \section{Bibliography}
    \begin{frame}{Bibliography}
        \begin{thebibliography}{99}
            \bibitem{Liouville1833} 
            Joseph Liouville, \textit{Sur la détermination des intégrales dont la valeur est algébrique}, 
            \textit{Journal de l'École Polytechnique}, vol. 14, 1833, pp. 93--123.

            \bibitem{Rosenlicht1972} 
            Maxwell Rosenlicht, \textit{Integration in finite terms}, 
            \textit{American Mathematical Monthly}, vol. 79, no. 9, 1972, pp. 963--972.

            \bibitem{Conrad2003} 
            Keith Conrad, \textit{Impossibility theorems for elementary integration}, 
            \textit{American Mathematical Monthly}, vol. 110, no. 5, 2003, pp. 459--462.
        \end{thebibliography}
    \end{frame}
\end{document}
